---
title: "ESM 244 Lab 1"
author: "Katelin Seeto"
date: "1/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# attach packages
library(tidyverse)
library(here)
library(tmap)
library(sf)

```

### Read in SF trees data

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))
```

### Basic wrangling reminders

Refresh some skills for data wrangling and summary statistics using functions from the `dplyr` package. 

Find the top 5 highest observations of trees by legal_status, do some wrangling, make a graph. 

Notes on `count()`:  
  * `dplyr` package with the job to find counts of observations based on one or more variables  
  * replaces need for `group_by` and n function and `summarise()` since it puts it in a nice table
  
Notes on other functions:  
  * `drop_na()` removes any rows that have na value for variable you specify  
  * `rename()` renames variable in df, remember new name on left, old name on right!   
    + rename(new = old)  
  * `select()` main job is to keep or exclude columns, can also be used to reorder but can be a hassle if many columns  
  * `relocate()` is better way to reorder columns, column name you put in will bring to first column  
  * `slice_max()` allows you to identify the rows with the highest values for a specified variables and then keeps the number of rows you want
    
  

```{r}
# Top 5 by legal_status
top_5_status <- sf_trees %>%
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n = 5)
  

# Example count by legal_status and species
# sf_trees %>% count(legal_status, species)
```


### Make a graph of top 5 observations by legal status 

Notes on ggplot:  
* Initial plotting shows all columns listed in alphabetical order
* Can convert to factor and use forcats package to reorder variables  
* We will reorder columns within `ggplot`, use `fct_reorder()` with (variable, variable to order by)

```{r}
ggplot(data= top_5_status, aes(x= fct_reorder(legal_status, tree_count), y= tree_count))+
  geom_col()+
  labs(x= "Legal Status",
       y= "Tree Count")+
  coord_flip()+
  theme_minimal()
```

### Few more data wrangling examples

Only want to keep observations (aka rows) for Blackwood Acacia trees 

Notes:  
* If you want to look for everything in a column that contains a string of text, like Blackwood Acacia can use `filter(str_detect(variable, "string of text"))`  
* Can use `ggplot` to map even though lat and long are just being viewed as numbers on an x,y coord system

```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, latitude, longitude)

ggplot(data= blackwood_acacia, aes(x= longitude, y= latitude))+
  geom_point()
```

### More wrangling examples: `tidyr:: separate()` and `unite()` functions  

Useful for combining or separating columns 

Notes:  
* `separate()` is broken down into this structure separate(column you want to separate, into = what you want to name them which is done by creating a vector with c, sep= what is the separator in "")  
* `unite()` is for combining columns, structure is ("name of new col you want to make', col's you want to unite, can use : to indicate from this col to this other col, sep= "what you want to separate them")

```{r}
# Example separate()
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep= "::")

# Example unite()

sf_trees_unite <- sf_trees %>% 
  unite("id_status", tree_id: legal_status, sep= "_cool_")
```

